---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# apodetect

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![R-CMD-check](https://github.com/selkamand/apodetect/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/selkamand/apodetect/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

> [!WARNING]\
> This package is in early development and not yet ready for use

**apodetect** identifies APOBEC-enriched samples and classifies the type of APOBEC identified based on positional biases (diffuse, focal, mixed) and most probable mutator enzyme (APOBEC3A or APOBEC3B).

Existing signature-powered approaches consistently miss APOBEC-induced tumours when contributing only low numbers of APOBEC. No method yet exists that describes the full complement of APOBEC-relevant features.

## Approach

**Detecting and quantifying the proportion of APOBEC enrichment**

1.  Count the number of mutations that match the common APOBEC-trinucleotide biases
    a.  [C\>G] TCA\>TGA; TCT\>TGT; TCC\>TGC (less common)
    b.  [C\>T] TCA\>TTA; TCT\>TTT; TCC\>TTC (less common)
2.  Count 4bp motifs contexts (helps differentiate APOBEC3A from APOBEC3B) (10.1038/s41467-024-46231-w).
    a.  A3A prefers YTCW motifs (where Y = C or T; W = A or T).
    b.  A3B prefers RTCW motifs (where R = G or A; W = A or T)
3.  Quantify if the ratios of TCA\>TGA:TCC\>TGC & TCA\>TTA:TCC\>TTC (if these deviate substantially from expected, it suggests dilution by other signatures).
4.  Quantify the bias of these types of mutations to the lagging strand (APOBEC mutations are consistently biased to lagging-strand)
5.  Quantify the bias of candidate mutations to hairpin stuctures (10.1126/science.aaw2872)
6.  Quantify the number of mutations at the 3' end of 3-nucleotide hairpins vs 4-nucleotide hairpins. A higher 3nt hairpin ratio suggests A3B. We may want to plot the See Supp Figure S2 of (10.1038/s41467-024-46231-w).

**Classifying APOBEC as diffuse, focal, or mixed**

1.  Annotate probable APOBEC mutations
2.  Identify sites of kataegis (mutation density)
3.  Quantify the proportion of APOBEC mutations that are part of regions of kataegis (clustered)
4.  Quantify the proportion of 'free' (non kataegis-proximal) APOBEC
5.  [todo] figure a sensible approach for checking if this 'free' APOBEC is actually uniformly distributed (min TMB is probably required) - maybe just having a decent proportion of non-kataegis proximal APOBEC is sufficient?
6.  If \> 80% of APOBEC mutations are clustered then class = 'focal'. If \< 10% of APOBEC mutations are clustered, then class = 'diffuse'. Anything else we assume is 'mixed'.

## Installation

You can install the development version of apodetect from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("selkamand/apodetect")
```

## Quick Start

Coming Soon

```{r example}
library(apodetect)
## basic example code
```
